D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-18 ootavacuous.tex
More minor edits
Here's a number of small changes to the text.  They are intended to
improve grammar, style, and consistency.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Some workers recommend avoiding OOTA by ordering prior relaxed
loads before subsequent relaxed
stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
but this can require real instructions be
executed~\cite[Section 7.1]{Maranget2012TutorialARMPower},
consuming real time and real electrical power to solve a strictly
theoretical problem.
This might have been acceptable in the 1960s of some of the authors'
youths, but it is now the year 2024.

@@

As long as \co{z} is zero, and all else being equal, changes in the
value of \co{y} will not cause a change in the value stored to \co{x}.
As a result, the semantic dependency from \co{y} to \co{x} exists only
in executions where \co{z} is nonzero,
which shows it is a property of the execution, not just of the source code.

@@

Suppose that in some execution of the previous example,
both \co{y} and \co{z} are zero.
Then changes to either \co{y} or \co{z} (again, all
else being equal) will not cause a change in the value stored to \co{x}.
In other words, in this execution there is no semantic dependency
from either \co{y} or \co{z} to \co{x}.
But there \emph{is} a semantic dependency from the \emph{pair}
\{\co{y},~\co{z}\} to \co{x},
because changes to both \co{y} and \co{z}
can cause the value stored in \co{x} to change.
This means that accurate definitions of sdep cannot always rely on
single-variable perturbations;
they must consider changes to multiple variables.
See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
for examples and additional discussion.

Since we can no longer regard sdep as always relating a single load to a store,
the notion of a cycle involving sdep appears problematic.
We are forced to change our definition of an OOTA cycle again;
we will say that an execution is an instance of OOTA if in that execution:
\begin{quote}
	There are stores
	$W_i$~($i = 0$, \ldots,~$m$),
	where $W_i$ semantically depends on loads
	$\{R_{i,j}\}$~($j = 0$, \ldots,~$n_i$),
	such that each $R_{i,j}$ reads from one of the $W_k$
	stores in a different thread.
\end{quote}
This makes OOTA more complicated than a simple cycle but
we will continue to refer to OOTA cycles out of habit.
Note that this new definition includes and generalizes the earlier
``cycle in sdep $\cup$ rfe'' definition.

@@

Consider the following:
\begin{quote}
\begin{verbatim}
x = y - z;
\end{verbatim}
\end{quote}
There appear to be semantic dependencies from \co{y} to \co{x} and from \co{z}
to \co{x}.
However, if the implementation somehow knows that \co{y} is
always equal to \co{z} at this point then there is no semantic dependency;
the implementation can act as if the statement were simply ``\co{x = 0;}''.
We leave aside the question of how the implementation would know this,
given that \co{y} and \co{z} cannot be updated simultaneously
and are subject to change at any time by
other threads (a point we will return to in
section~\ref{sec:Global Optimization Can Destroy Dependencies}).
% The usual way in the Linux kernel is via a union, but it is probably
% best to ignore that possibility.

@@

Suppose we take the view that the previous example involves only one
store.
This opens up the door to greater complexity:
\begin{quote}
\begin{verbatim}
if (x > 0) {
    L1: y = 42;
} else {
    y = 53;
    y = 42;
}
\end{verbatim}
\end{quote}
Consider an execution in which \co{x} is greater than zero, so the
statement labeled \co{L1} runs.
Is it semantically dependent on \co{x}?
The answer isn't immediately clear.
If the other arm of the \co{if} is taken then the same store of 42 to
\co{y} occurs, but 53 is written before it.
Which of these two stores should be compared with the store in \co{L1}?

@@

We have seen several examples showing that semantic dependencies may vary
according to the execution and even the implementation.
This raises some questions, of which the first is:
What exactly is an execution?

@@

When a given computer executes the machine code in a file, the outcome
has historically been much better defined than the executions of the
C++ abstract machine.
The hardware's behavior is typically specified with great precision by
the designer or manufacturer, and there are formal, executable memory models
describing exactly what patterns of loads and stores can occur.
Thus, leaving aside questions of asynchronous interrupts and system
calls, the behavior of a CPU executing a particular thread within a
program is entirely determined by the values obtained by the various
load-from-memory instructions.

@@

Consider this code, in which the final values of \co{r1} and~\co{r2}
are observable:
\begin{quote}
\begin{verbatim}
 1  int r1 = (x != 0);
 2  int r2 = (y != 0);
 3  z = (r1 == r2);
\end{verbatim}
\end{quote}
It is clear that the store to \co{z} semantically depends on the load
from \co{y}, because the value of \co{z} will change whenever \co{y}
changes between zero and nonzero (all else being equal).

@@

To express this idea in more formal terms,
and to explain what we mean by ``quasi-volatile'' object accesses,
we augment the
requirements for a hardware execution $H$ to realize an abstract
execution $A$.
Each realization must include a map from the set of accesses of
volatile objects in $A$ to the set of instructions in $H$ that access
these objects, having the following properties:
\begin{itemize}
\item	The map connects accesses of the same type (loads to loads
	and stores to stores) and to the same object.
\item	The map connects accesses in a thread of $A$ to accesses in
	the corresponding thread of $H$.
\item	The map is value-preserving: The value of an access in $A$ must be
	the same as the value of the access it maps to in $H$.
\item	In benign environments the map must preserve the rf relation.
	That is, if volatile load $R$ in $A$ reads from store $W$ then
	then the instruction it maps to in $H$ must read from the
	instruction that $W$ maps to.
\item	The map is order-preserving: Two accesses in the same thread
	of $A$ must map to accesses occurring in the same order in $H$.
	(In other words, the compiler may not reorder accesses
	to volatile objects.)
\item	The map is onto: For every access in $H$ to a volatile object
	there must be an access in $A$ that maps to it.
	(In other words, the compiler may not invent accesses to
	volatile objects.)
\item	The map is one-to-one: Different accesses in $A$ map to
	different accesses in $H$.
	(In other words, the compiler may not merge accesses to
	volatile objects.)
\item	The map is total: Every access in $A$ to a volatile object
	maps to some access in $H$.
	(In other words, the compiler may not omit accesses to
	volatile objects.)
\end{itemize}
Most of these are direct consequences of the fact that volatile-object
accesses are considered to be a form of I/O when the program runs
in a nonbenign environment.
But to be clear, these requirements apply to all environments.

@@

Dependencies between machine instructions can be understood in terms
of the flow of information within a CPU.
Each instruction has a set of inputs and is the source of a set of outputs,
some of which flow to the inputs for later instructions.
The inputs determine what an instruction will do.
A few examples:
\begin{itemize}
\item	An \co{add} instruction would typically have two inputs
	(the register values to be added together) and two outputs
	(the sum to be stored in a general-purpose register and some
	condition-code bits---e.g., Zero, Carry, and Overflow---to be
	stored in a status register).
\item	A conditional move instruction would have as inputs the
	condition-code bits to test, the source register value, and the
	target register value; the output would be the value to be stored
	in the target register.
\item	A conditional or computed branch instruction would have as inputs
	the condition-code bits to test, the address of the following
	instruction (to be used if the condition is false) and the
	destination address (to be used if the condition is true).
	The output is the new address to be written to the
	instruction-pointer register.
\item	A memory-load instruction's input is the address to load from,
	and its output is the value obtained by the load, to be stored
	in a register.
\item	A memory-store instruction's inputs are the value to store and
	the address at which to store it; there are no outputs.
\end{itemize}
Note: Hardware dependency analysis considers only the information
flowing \emph{within} a CPU, not information flowing between the
CPU and memory, which is handled separately as part of the action
taken by an instruction.

@@

A CPU may start executing an instruction speculatively, but at some
point it must \emph{commit} to a decision either to definitely execute the
instruction's action with some collection of well defined inputs and outputs
or else to abandon it.
Thanks to the law of cause and~effect, a CPU is not able to commit an
instruction or its outputs until the relevant inputs' sources have committed.
The reason is obvious: An input is subject to change at any time until
its source commits to its value,
and the instruction and its outputs can't commit until the CPU has fully
determined what it will do and what they will be.
(On the other hand, the CPU need not wait for inputs that won't affect
the instruction's results.
For instance, a conditional move instruction wouldn't need to wait for
the source register input to be determined once it knew that the
condition was definitely false.
And, although no existing architectures do this, there is nothing
in principle to prevent a CPU from committing a \co{multiply}
instruction as soon as either one of its inputs' sources has committed to the
value zero.)

@@

A similarly straightforward mechanism for ordering is conditional or
computed branches.
An instruction executing after such a branch cannot commit
until the CPU has committed to whether the branch will be taken and
if taken, where it will branch to (that is, the output value it will
send to the instruction-pointer register);
until then the CPU cannot know whether the later instruction should
be executed at all.
Therefore instructions following a conditional or computed branch
instruction in an execution must commit after the branch, and hence
after the source for the branch's condition or destination input.
(There could theoretically be an exception for a branch that
conditionally jumps to the immediately following instruction.
We can ignore this possibility by treating such branches as no-ops.)

@@

With the contents of the last few sections under our belts, we can
formulate the Fundamental Property that we would like all semantic
dependencies to satisfy.
Note that this formulation makes sense only for implementations in
which all atomic objects are treated as quasi volatile.
\begin{quote}
Let $W$ be a store which, in some abstract execution $A$, semantically
depends on loads $\{R_0$, \ldots,~$R_n\}$, and suppose that $W$ is not
omitted in some hardware execution $H$ realizing $A$.
Then for some $i$, load $R_i$ is not omitted and the hardware
instruction it maps to is ordered before the instruction $W$ maps to.
\end{quote}
It's quite straightforward to show that for any implementation in which
semantic dependencies satisfy the Fundamental Property, no OOTA cycle
has a nontrivial realization.

@@

Let $A$ be an abstract execution of some program $P$ containing a
thread $T$, and let $H$ be a hardware execution realizing $A$.
Let $W$ be an store in $T$ to an atomic object and let $\{R_i \mid i =
0, \ldots, n\}$ be a set of loads in $T$ from atomic objects on which
$W$ might or might not depend.
We can dispose of one case immediately: If $W$ is omitted in $H$ then
the issue of semantic dependency is moot.
You can give either answer since the question is basically meaningless.
Therefore we'll assume that $W$ is not omitted.
Then:
\begin{quote}
There is a semantic dependency from $\{R_i\}$ to $W$ in $A$ and $H$,
relative to the compiler used to produce $H$, if there is another
abstract execution $B$ realized by hardware execution $G$ under the
same compiler that together \emph{witness the semantic dependency}.
\end{quote}

To be a proper witness, $B$ must be an execution of some program $Q$,
not necessarily the same as $P$ but which contains the same thread $T$.
The thread should start out with the same initial state in $A$ and
$B$, and all loads in $A$ coming before any of the $R_i$ should obtain
the same value as they do in $B$ (this is part of our interpretation
of ``all other things being equal'').
It follows that the two abstract executions of $T$ will be identical
up to the first of the $R_i$ loads.

Let $W'$ and $\{R'_i\}$ be the accesses in $H$ that $W$ and the
non-omitted $\{R_i\}$ loads map to.
We then require that the hardware executions of $T$ in $H$ and $G$ be
identical for an initial period lasting up to the first of the $R'_i$.
Following this initial period there will be a common period, during
which $H$ and $G$ execute the same machine instructions but do not
necessarily compute the same values.
This common period ends when one of the hardware executions
takes a conditional branch that the other doesn't, or when a computed
branch leads to different addresses in the two executions, or when $T$
ends, whichever comes first.
Past this point $H$ and $G$ diverge and are no longer directly
comparable, as they execute different instructions.
Our third requirement for being a witness is that each load in the
common portion either must obtain the same value in $H$ and $G$, or
must itself be one of the $R'_i$ loads, or must be ordered in $H$
after one of the $R'_i$ loads (this is the remaining part of our
interpretation of ``all other things being equal''.)

Finally, we need to determine an instruction $X'$ in $G$ that
corresponds to $W'$.
If $W'$ is part of the initial or common portion of $H$ this is no
problem; we can take $X'$ to be $W'$ itself.
But if $W'$ is in the divergent portion of $H$ then things aren't so
simple.
The choice is somewhat arbitrary, and so we will fall back on the
earlier proposal of matching up stores by the order they occur.
Let \co{y} be the atomic object that $W'$ stores to, and suppose $W'$
is the $N$th store to \co{y} within the divergent portion of $H$.
Then $X'$ will be the $N$th store to \co{y} in the divergent portion
of $G$, if such a store exists.
Our last requirement for being a witness to a semantic dependency
is that $X'$ act differently from $W'$: it doesn't exist, it stores
a different value, or it stores to an object other than \co{y}.

@@

As promised earlier, we now consider implementations whose compilers
may use global analysis.
In order to obtain the desired results we have to require that these
compilers treat all atomic objects as volatile.
Equivalently, the machine code generated by such a compiler must be
the same for a given program as for a ``volatilized'' form of the
program in which all atomic objects are defined to be volatile.

In this context our definition of semantic dependency is essentially
the same as before.
Since we can no longer expect the machine code for a thread to be the
same regardless of the program it belongs to, the program $Q$ in the
earlier definition (of which $B$ and $G$ are executions) must now be
$P$ or its volatilized form.
However, we do now allow the possibility that the executions $B$ and
$G$ take place in a nonbenign environment.
Aside from these minor adjustments, the definition remains unchanged.

@@

Third case: $W'$ lies in the divergent portion of $H$.
This happens when $W'$ comes after the conditional or computed branch
which marks the end of the common period by going different ways in
$H$ and $G$.
Just as in the previous case, since the branch behaves differently in
the two executions it must be ordered after one of the $R'_i$ loads.
And then so must $W'$, because any instruction following a conditional
or computed branch instruction is ordered after the branch.
QED.

@@

A drawback of the definition is that it is only relative to a specific
compiler or implementation.
This may strike some people as wishy-washy and as avoiding the real
problem, in that a given piece of code should either contain or not
contain a semantic dependency independent of the implementation used to
run it.

We can address this drawback by defining an \emph{absolute semantic
dependency} as one that is present relative to any valid C++
implementation, past, present, or future.
Of course this notion has its own problems, including that it is
extremely nonconstructive and impossible to apply in practice.
However it may be the best we can do with our current understanding
of computing systems.

There is one thing we can definitely state:
Programs produced by an implementation of the right sort will never
exhibit absolute OOTA (that is, an OOTA cycle in which all the
semantic dependencies are absolute).
This is simply because a semantic dependency that is absolute is {\it
a fortiori\/} a semantic dependency relative to the compiler in use.

@@

To see why, recall that a semantic dependency must have a witness,
another execution in which the store acts differently.
But this transformed program has no other executions; every time it
runs it will write 42 to both \co{x} and \co{y}.
(Keep in mind also that since the atomic objects are not treated as
volatile, they are not subject to unspecified interference when the
program runs in a nonbenign environment.)

@@

Consider this simple example:
\begin{quote}
\begin{verbatim}
x = (int) &x;
\end{verbatim}
\end{quote}
Even though the example contains no loads at all, it may store
different values when running in different programs because the
variable \co{x} may be allocated at differing addresses in those
programs.
According to our definition, this could count as a degenerate OOTA
cycle of length one, in which the store is semantically dependent on
an empty set of loads!

@@

However, a single-thread analysis compiler can translate this into the
machine code equivalent of:
\begin{quote}
\begin{verbatim}
int r1 = (x != 0);
int r2;
int r3 = (x != 0);
int r4 = y;
if (r1 != r3) {
   r2 = r4;
   z = 1;
} else {
   r2 = r1;
   z = (r2 == r4);
}
\end{verbatim}
\end{quote}
In effect, the \co{r2} load is merged with the \co{r1} load in
executions where \co{r1} is equal to \co{r3} or to \co{r4}, and it is
merged with the \co{r3} load in other executions.

@@

\begin{listing}
\begin{verbatim}
 1 atomic<int> x, y;
 2
 3 void thread1() {
 4   int r1 = x.load(memory_order::relaxed);
 5   y.store(r1, memory_order::relaxed);
 6 }
\end{verbatim}
\caption{Non-Volatile Accesses and Dependencies}
\label{lst:Non-Volatile Accesses and Dependencies}
\end{listing}

@@

\begin{listing}
\begin{verbatim}
 1 atomic<volatile int> x, y;
 2
 3 void thread1() {
 4   int r1 = x.load(memory_order::relaxed);
 5   y.store(r1, memory_order::relaxed);
 6 }
\end{verbatim}
\caption{Volatile Accesses and Dependencies}
\label{lst:Volatile Accesses and Dependencies}
\end{listing}

>>>>

Some workers recommend avoiding OOTA by forcing prior relaxed
loads to be ordered before subsequent relaxed
stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
but this can require real instructions be
executed~\cite[Section 7.1]{Maranget2012TutorialARMPower},
consuming real time and real electrical power to solve a strictly
theoretical problem.
This might have been acceptable in the 1960s of some of the authors'
youths, but it is now the year 2024.

@@

As long as \co{z} is zero, changes in the
value of \co{y} will not cause a change in the value stored to \co{x}.
As a result, the semantic dependency from \co{y} to \co{x} exists only
in executions where \co{z} is nonzero,
which shows it is a property of the execution, not just of the source code.

@@

Suppose that in some execution of the previous example,
both \co{y} and \co{z} are zero.
Then changes to either \co{y} or \co{z}
will not cause a change in the value stored to \co{x}.
In other words, in this execution there is no semantic dependency
from either \co{y} or \co{z} to \co{x}.
But there \emph{is} a semantic dependency from the \emph{pair}
\{\co{y},~\co{z}\} to \co{x},
because changes to both \co{y} and \co{z}
can cause the value stored in \co{x} to change.
This means that accurate definitions of sdep cannot always rely on
single-variable perturbations;
they must consider changes to multiple variables.
See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
for examples and additional discussion.

Since we can no longer regard sdep as always relating a single load to a store,
the notion of a cycle involving sdep appears problematic.
We are forced to change our definition of an OOTA cycle again;
we will say that an execution is an instance of OOTA if in that execution:
\begin{quote}
	There are stores $W_0,$ \ldots,~$W_m$,
	where each $W_i$ semantically depends on a set of loads
	$\{R_{i,0},$ \ldots,~$R_{i,n_i}\}$,
	such that each $R_{i,j}$ reads from one of the $W_k$
	stores in a different thread.
\end{quote}
This makes OOTA more complicated than a simple cycle but
we will continue to refer to OOTA cycles out of habit.
Note that this new definition includes and generalizes the earlier
``cycle in sdep $\cup$ rfe'' definition.

@@

Consider the following:
\begin{quote}
\begin{verbatim}
x = y - z;
\end{verbatim}
\end{quote}
There appear to be semantic dependencies from \co{y} to \co{x} and from \co{z}
to \co{x}.
However, if the implementation somehow knows that \co{y} is
always equal to \co{z} at this point then there is no semantic dependency;
the implementation can act as if the statement were simply ``\co{x = 0}''.
We leave aside the question of how the implementation would know this,
given that \co{y} and \co{z} cannot be updated simultaneously
and are subject to change at any time by
other threads (a point we will return to in
section~\ref{sec:Global Optimization Can Destroy Dependencies}).
% The usual way in the Linux kernel is via a union, but it is probably
% best to ignore that possibility.

@@

Suppose we take the view that the previous example involves only one
store.
This opens up the door to greater complexity:
\begin{quote}
\begin{verbatim}
if (x > 0) {
    L1: y = 42;
} else {
    y = 53;
    y = 42;
}
\end{verbatim}
\end{quote}
Consider an execution in which \co{x} is greater than zero, so the
statement labeled \co{L1} runs.
Is it semantically dependent on \co{x}?
The answer isn't immediately clear.
If the other arm of the \co{if} is taken then a store of the same value 42 to
\co{y} occurs, but 53 is written before it.
Which of these two stores should be compared with the store in \co{L1}?

@@

We have seen several examples showing that semantic dependencies may vary
according to the execution and even the implementation.
This raises several questions, of which the first is:
What exactly is an execution?

@@

The outcome when a given computer executes the machine code in a file
has historically been much better defined than the executions of the
C++ abstract machine.
The hardware's behavior is typically specified with great precision by
the designer or manufacturer, and there are formal, executable memory models
describing exactly what patterns of loads and stores can occur.
Thus, leaving aside questions of asynchronous interrupts and system
calls, the behavior of a CPU executing a particular thread within a
program is entirely determined by the values obtained by the various
memory-load instructions.
(We count read-modify-write instructions as both memory loads and
memory stores.)

@@

Consider this code, in which the final values of \co{r1} and~\co{r2}
are observable:
\begin{quote}
\begin{verbatim}
   int r1 = (x != 0);
   int r2 = (y != 0);
   z = (r1 == r2);
\end{verbatim}
\end{quote}
It is clear that the store to \co{z} semantically depends on the load
from \co{y}, because the value of \co{z} will change whenever \co{y}
changes between zero and nonzero (all else being equal).

@@

To express this idea in more formal terms,
and to explain what we mean by ``quasi-volatile'' object accesses,
we augment the
requirements for a hardware execution $H$ to realize an abstract
execution $A$.
Each realization must include a map from the set of accesses of
volatile objects in $A$ to the set of instructions in $H$ that access
these objects, having the following properties:
\begin{itemize}
\item	The map connects accesses of the same type (loads to loads
	and stores to stores) and to the same object.
\item	The map connects accesses in a thread of $A$ to accesses in
	the corresponding thread of $H$.
\item	The map is value-preserving: The value of an access in $A$ must be
	the same as the value of the access it maps to in $H$.
\item	In benign environments the map must preserve the rf relation.
	That is, if volatile load $R$ in $A$ reads from store $W$ then
	then the instruction it maps to in $H$ must read from the
	instruction that $W$ maps to.
\item	The map is order-preserving: Two accesses in the same thread
	of $A$ must map to accesses occurring in the same order in $H$.
	(In other words, the compiler may not reorder accesses
	to volatile objects.)
\item	The map is onto: For every access in $H$ to a volatile object
	there must be an access in $A$ that maps to it.
	(In other words, the compiler may not invent accesses to
	volatile objects.)
\item	The map is one-to-one: Different accesses in $A$ map to
	different accesses in $H$.
	(In other words, the compiler may not merge accesses to
	volatile objects.)
\item	The map is total: Every access in $A$ to a volatile object
	maps to some access in $H$.
	(In other words, the compiler may not omit accesses to
	volatile objects.)
\end{itemize}
Most of these are direct consequences of the fact that volatile-object
accesses are considered to be a form of I/O when the program runs
in a nonbenign environment.
But to be clear, these requirements apply in all environments.

@@

Dependencies between machine instructions can be understood in terms
of the flow of information within a CPU.
Each instruction has a set of inputs and is the source of a set of outputs,
some of which flow to the inputs for later instructions.
The inputs determine what an instruction will do.
A few examples:
\begin{itemize}
\item	An \co{add} instruction would typically have two inputs
	(the register values to be added together) and two outputs
	(the sum to be stored in a general-purpose register and some
	condition-code bits---e.g., Zero, Carry, and Overflow---to be
	stored in a status register).
\item	A conditional-move instruction would have as inputs the
	condition-code bits to test, the source register value, and the
	target register value; the output would be the value to be stored
	in the target register.
\item	A conditional- or computed-branch instruction would have as inputs
	the condition-code bits to test, the address of the following
	instruction (to be used if the condition is false) and the
	destination address (to be used if the condition is true).
	The output is the new address to be written to the
	instruction-pointer register.
\item	A memory-load instruction's input is the address to load from,
	and its output is the value obtained by the load, to be stored
	in a register.
\item	A memory-store instruction's inputs are the value to store and
	the address at which to store it; there are no outputs.
\end{itemize}
Note: Hardware dependency analysis considers only the information
flowing \emph{within} a CPU, not information flowing between the
CPU and memory, which is handled separately as part of the action
taken by an instruction.

@@

A CPU may start executing an instruction speculatively, but at some
point it must \emph{commit} to a decision either to definitely execute the
instruction's action with some collection of well defined inputs and outputs
or else to abandon it.
Thanks to the law of cause and~effect, a CPU is not able to commit an
instruction or its outputs until the relevant inputs' sources have committed.
The reason is obvious: An input is subject to change at any time until
its source commits to its value,
and the instruction and its outputs can't commit until the CPU has fully
determined what it will do and what they will be.
(On the other hand, the CPU need not wait for inputs that won't affect
the instruction's results.
For instance, a conditional-move instruction wouldn't need to wait for
the source register input to be determined once it knew that the
condition was definitely false.
And, although no existing architectures do this, there is nothing
in principle to prevent a CPU from committing a \co{multiply}
instruction as soon as either one of its inputs' sources has committed to the
value zero.)

@@

A similarly straightforward mechanism for ordering is conditional or
computed branches.
An instruction executing after such a branch cannot commit
until the CPU has committed to whether the branch will be taken and
if taken, where it will branch to (that is, the output value it will
send to the instruction-pointer register);
until then the CPU cannot know whether the later instruction should
be executed at all.
Therefore instructions following a conditional- or computed-branch
instruction in an execution must commit after the branch, and hence
after the source for the branch's condition or destination input.
(There could theoretically be an exception for a branch that
conditionally jumps to the immediately following instruction.
We can ignore this possibility by treating such branches as no-ops.)

@@

With the contents of the last few sections under our belts, we can
formulate the Fundamental Property that we would like all semantic
dependencies to satisfy.
Note that this formulation makes sense only for implementations in
which all atomic objects are treated as volatile or quasi volatile.
\begin{quote}
Let $W$ be a store which semantically depends on loads
$\{R_0$, \ldots,~$R_n\}$ in some abstract execution $A$,
and suppose that $W$ is not
omitted in some hardware execution $H$ realizing $A$.
Then for some $i$, load $R_i$ is not omitted and the hardware
instruction it maps to is ordered before the instruction $W$ maps to.
\end{quote}
It's quite straightforward to show that for any implementation in which
semantic dependencies satisfy the Fundamental Property, no OOTA cycle
has a nontrivial realization.

@@

Let $A$ be an abstract execution of some program $P$ containing a
thread $T$, and let $H$ be a hardware execution realizing $A$.
Let $W$ in $T$ be a store to an atomic object, and let $\{R_0,$
\ldots,~$R_n\}$ in $T$ be a set of loads from atomic objects on which
$W$ might or might not depend.
We can dispose of one case immediately: If $W$ is omitted in $H$ then
the issue of semantic dependency is moot.
You can give either answer since the question is basically meaningless.
Therefore we'll assume that $W$ is not omitted.
Then:
\begin{quote}
There is a semantic dependency from $\{R_i\}$ to $W$ in $A$ and $H$,
relative to the compiler used to produce $H$, if there is another
abstract execution $B$ realized by hardware execution $G$ under the
same compiler that together \emph{witness the semantic dependency}.
\end{quote}

To be a proper witness, $B$ must be an execution of some program $Q$,
not necessarily the same as $P$ but which contains the same thread $T$.
The thread should start out with the same initial state in $A$ and
$B$, and all loads in $A$ coming before any of the $R_i$ should obtain
the same value as they do in $B$ (this is part of our interpretation
of ``all else being equal'').
It follows that the two abstract executions of $T$ will be identical
up to the first of the $R_i$ loads.

Let $W'$ and $\{R'_i\}$ be the accesses in $H$ that $W$ and the
non-omitted $\{R_i\}$ loads map to.
We then require that the hardware executions of $T$ in $H$ and $G$ be
identical for an initial period lasting up to the first of the $R'_i$.
Following this initial period there will be a common period, during
which $H$ and $G$ execute the same machine instructions but do not
necessarily compute the same values.
This common period ends when one of the hardware executions
takes a conditional branch that the other doesn't, or when a computed
branch leads to different addresses in the two executions, or when $T$
ends, whichever comes first.
Past this point $H$ and $G$ diverge and are no longer directly
comparable, as they execute different instructions.
Our third requirement for being a witness is that each load in the
common portion either must obtain the same value in $H$ and $G$, or
must itself be one of the $R'_i$ loads, or must be ordered in $H$
after one of the $R'_i$ loads (this is the remaining part of our
interpretation of ``all else being equal''.)

Finally, we need to determine an instruction $X'$ in $G$ that
corresponds to $W'$.
If $W'$ is part of the initial or common portion of $H$ this is no
problem; we can take $X'$ to be $W'$ itself.
But if $W'$ is in the divergent portion of $H$ then things aren't so
simple.
The choice is somewhat arbitrary, and so we will fall back on the
earlier proposal of matching up stores by the order they occur.
Let $y$ be the atomic object that $W'$ stores to, and suppose $W'$
is the $N$th store to $y$ within the divergent portion of $H$.
Then $X'$ will be the $N$th store to $y$ in the divergent portion
of $G$, if such a store exists.
Our last requirement for being a witness to a semantic dependency
is that $X'$ act differently from $W'$: it doesn't exist, it stores
a different value, or it stores to an object other than $y$.

@@

As promised earlier, we now consider implementations whose compilers
may use global analysis.
In order to obtain the desired results we have to require that these
compilers treat all atomic objects as volatile.
Equivalently, the machine code generated by such a compiler must be
the same for a given program as for a ``volatilized'' form of the
program in which all the atomic objects are defined to be volatile.

In this context our definition of semantic dependency is essentially
the same as before.
Since we can no longer expect the machine code for a thread to be the
same regardless of the program it belongs to, the program $Q$ in the
earlier definition (of which $B$ and $G$ are executions) must be
$P$ or its volatilized form.
However, we do now allow the possibility that the executions $B$ and
$G$ take place in a nonbenign environment.
Aside from these minor adjustments, the definition remains unchanged.

@@

Third case: $W'$ lies in the divergent portion of $H$.
This happens when $W'$ comes after the conditional or computed branch
which marks the end of the common period by going different ways in
$H$ and $G$.
Just as in the previous case, since the branch behaves differently in
the two executions it must be ordered after one of the $R'_i$ loads.
And then so must $W'$, because any instruction following a conditional-
or computed-branch instruction is ordered after that branch.
QED.

@@

A drawback of the definition is that it is only relative to a specific
compiler or implementation.
This may strike some people as wishy-washy and as avoiding the real
problem, in that a given piece of code should either contain or not
contain a semantic dependency, independent of the implementation used to
run it.

We can address this drawback by defining an \emph{absolute semantic
dependency} as one that is present relative to any valid loose C++
implementation, past, present, or future, real or imagined.
Of course this notion has its own problems, including that it is
extremely nonconstructive and impossible to apply in practice.
However it may be the best we can do with our current understanding
of computing systems.

There is one thing we can definitely state:
Programs produced by an implementation of the right sort will never
exhibit absolute OOTA (that is, an OOTA cycle in which all the
semantic dependencies are absolute).
This is simply because an absolute semantic dependency is {\it
a fortiori\/} a semantic dependency relative to the compiler in use.

@@

To see why, recall that a semantic dependency must have a witness,
another execution in which the store acts differently.
But this transformed program has no other executions; every time it
runs it will store 42 to both \co{x} and \co{y}.
(Keep in mind also that since the atomic objects are not treated as
volatile, they are not subject to unspecified interference when the
program runs in a nonbenign environment.)

@@

Consider this simple example:
\begin{quote}
\begin{verbatim}
x = (int) &x;
\end{verbatim}
\end{quote}
Even though the example contains no loads at all, it may store
different values when running in different programs because the
object \co{x} may be allocated at differing addresses in those
programs.
According to our definition, this could count as a degenerate OOTA
cycle of length one, in which the store is semantically dependent on
an empty set of loads!

@@

However, a single-thread analysis compiler can translate this into the
machine-code equivalent of:
\begin{quote}
\begin{verbatim}
int r1 = (x != 0);
int r2;
int r3 = (x != 0);
int r4 = y;
if (r1 != r3) {
   r2 = r4;
   z = 1;
} else {
   r2 = r1;
   z = (r2 == r4);
}
\end{verbatim}
\end{quote}
In effect, the \co{r2} load is merged with the \co{r1} load in
executions where \co{r1} is equal to \co{r3} or to \co{r4}, and it is
merged with the \co{r3} load in other executions.

@@

\begin{listing}
\begin{verbatim}
 1 atomic<int> x, y;
 2
 3 void thread1() {
 4   int r1 = x.load(memory_order_relaxed);
 5   y.store(r1, memory_order_relaxed);
 6 }
\end{verbatim}
\caption{Non-Volatile Accesses and Dependencies}
\label{lst:Non-Volatile Accesses and Dependencies}
\end{listing}

@@

\begin{listing}
\begin{verbatim}
 1 atomic<volatile int> x, y;
 2
 3 void thread1() {
 4   int r1 = x.load(memory_order_relaxed);
 5   y.store(r1, memory_order_relaxed);
 6 }
\end{verbatim}
\caption{Volatile Accesses and Dependencies}
\label{lst:Volatile Accesses and Dependencies}
\end{listing}
